---
title: "Week 3: Visualizing Uncertainty and Change"
description: "Using ggplot2, margins of error, and ACS data to study uncertainty over time."
date: 2026-09-11
---

## What We Covered

This week connected data visualization with uncertainty in American Community Survey estimates. We practiced using `ggplot2`, calculating margins of error for derived values, comparing ACS estimates across time, and creating reusable R functions.

The examples below use New York county data and the ACS 2023 5-year survey. Most code chunks are marked `eval: false` so this study document can render without repeatedly downloading Census data. Remove that line when you want to run a chunk yourself.

## Setup

Install packages only once:

```{r}
#| eval: false

install.packages(c("tidyverse", "tidycensus", "gt"))
```

Load packages whenever you use them:

```{r}
#| eval: false

library(tidyverse)
library(tidycensus)
library(gt)
```

## The Grammar of Graphics

`ggplot2` is based on the idea that a plot is created by connecting variables in a data frame to visual properties.

| Data variable | Plot feature |
|---|---|
| Estimate | Position on an axis |
| Population | Position on an axis |
| Reliability | Color |
| Margin of error | Size or error bars |

This idea is called the **Grammar of Graphics**. A plot is built from separate parts that work together.

## Creating ACS Data

The first step is to load the packages and download the data. `B17001_002` represents the number of people below 100% of the poverty level.

```{r}
#| eval: false

library(tidyverse)
library(tidycensus)

my_variable <- "B17001_002"
my_state <- "NY"

county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

The returned data is in tidy, or long, format. Important columns include:

- `GEOID`: geographic identifier
- `NAME`: county name
- `estimate`: Census estimate
- `moe`: margin of error

## Calculating a Relative Margin of Error

To compare uncertainty across counties, calculate the margin of error as a percentage of the estimate:

```{r}
#| eval: false

county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

A county with an MOE of 500 may seem more uncertain than one with an MOE of 100, but that depends on the size of the estimates. The relative margin of error makes the comparison more meaningful.

## Creating Reliability Categories

```{r}
#| eval: false

county_data <- county_data %>%
  mutate(
    reliability = case_when(
      moe_pct < 5  ~ "High confidence",
      moe_pct < 10 ~ "Moderate",
      TRUE         ~ "Low confidence"
    )
  )
```

`case_when()` checks each condition from top to bottom. The first condition that is `TRUE` determines the category. `TRUE ~ "Low confidence"` catches all remaining values.

## The Basic Parts of a ggplot

A basic plot usually contains:

1. The data
2. Aesthetic mappings
3. A geometric object

```{r}
#| eval: false

ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point()
```

The data frame must be named correctly. This is only a placeholder and will cause an error if no object named `data` exists:

```{r}
#| eval: false

ggplot(data, aes(x = population, y = moe_pct)) +
  geom_point()
```

Use the actual data frame and column names instead:

```{r}
#| eval: false

ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point()
```

In the earlier wide population data, the equivalent plot would use `county_wide` and `popE`:

```{r}
#| eval: false

ggplot(county_wide, aes(x = popE, y = moe_pct)) +
  geom_point()
```

## Mapping Versus Decorating

If a visual property depends on a variable, place it inside `aes()`:

```{r}
#| eval: false

ggplot(county_data, aes(
  x = estimate,
  y = moe_pct,
  color = reliability
)) +
  geom_point()
```

If every point should have one fixed color, place the color outside `aes()`:

```{r}
#| eval: false

ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point(color = "blue")
```

These two forms mean different things:

```r
aes(color = reliability)       # maps a variable
geom_point(color = "blue")     # sets one fixed color
```

Writing `aes(color = "blue")` maps the constant text value `"blue"` to color. It is not the normal way to set a fixed color.

## Common Geometries

| Question | Useful geometry |
|---|---|
| How is one variable distributed? | `geom_histogram()` |
| How are two variables related? | `geom_point()` |
| How do groups compare? | `geom_boxplot()` or `geom_col()` |
| How does something change over time? | `geom_line()` |
| How uncertain is an estimate? | `geom_errorbar()` |

## Error Bars

An error bar can show an estimate together with its margin of error:

```{r}
#| eval: false

county_data %>%
  ggplot(aes(x = NAME, y = estimate)) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = estimate - moe,
      ymax = estimate + moe
    )
  )
```

This first version may be difficult to read because it includes every county. We can keep the 15 counties with the largest relative margins of error:

```{r}
#| eval: false

county_data %>%
  arrange(desc(moe_pct)) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = NAME, y = estimate)) +
  geom_col() +
  geom_errorbar(
    aes(ymin = estimate - moe, ymax = estimate + moe)
  )
```

County names are easier to read when the chart is flipped:

```{r}
#| eval: false

county_data %>%
  arrange(desc(moe_pct)) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = NAME, y = estimate)) +
  geom_col() +
  geom_errorbar(
    aes(ymin = estimate - moe, ymax = estimate + moe),
    width = 0.3
  ) +
  coord_flip()
```

To order the bars by estimate rather than alphabetically, use `reorder()`:

```{r}
#| eval: false

county_data %>%
  arrange(desc(moe_pct)) %>%
  mutate(NAME = str_remove(NAME, " County, New York")) %>%
  slice_head(n = 15) %>%
  ggplot(aes(x = reorder(NAME, estimate), y = estimate)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(
    aes(ymin = estimate - moe, ymax = estimate + moe),
    width = 0.3
  ) +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Counties with the largest margins of error",
    subtitle = "People below 100% of the poverty level, ACS 2023 5-year estimates",
    x = NULL,
    y = "Population below poverty level"
  )
```

## Facets and Small Multiple Plots

A facet creates a separate panel for each category of a variable:

```{r}
#| eval: false

ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point() +
  facet_wrap(~reliability)
```

## Themes and Labels

A theme changes the appearance of a plot. It does not change the data.

```{r}
#| eval: false

ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point() +
  theme_minimal() +
  labs(
    title = "Estimate and Relative Margin of Error",
    x = "Estimate",
    y = "Margin of error percentage"
  )
```

## Derived Uncertainty

When we calculate a new value from Census estimates, the new value also has uncertainty.

For example, if we add the number of people below poverty across several counties, we need the margin of error for the combined estimate. We should not simply add the individual MOEs.

```{r}
#| eval: false

region <- county_data %>%
  filter(NAME %in% c(
    "Hamilton County, New York",
    "Essex County, New York",
    "Greene County, New York"
  ))

region_total <- sum(region$estimate)
region_total_moe <- moe_sum(
  region$moe,
  region$estimate
)

region_total
region_total_moe
```

The function is used as follows:

```r
moe_sum(moe, estimate)
```

The margin of error for the total is not found by adding the individual MOEs together.

We can also calculate a margin of error for each reliability group:

```{r}
#| eval: false

county_data %>%
  group_by(reliability) %>%
  summarize(
    total = sum(estimate),
    total_moe = moe_sum(moe, estimate)
  )
```

## Proportions and `moe_prop()`

Suppose we want the proportion of people below poverty out of a total population. Both the numerator and denominator have estimates and margins of error.

```{r}
#| eval: false

poverty_share <- get_acs(
  geography = "county",
  variables = c(
    below_poverty = "B07012_002",
    total = "B07012_001"
  ),
  state = "NY",
  year = 2023,
  survey = "acs5",
  output = "wide"
) %>%
  mutate(
    share = below_povertyE / totalE,
    share_moe = moe_prop(
      below_povertyE,
      totalE,
      below_povertyM,
      totalM
    ),
    share_pct = share * 100,
    share_moe_pct = share_moe * 100
  )
```

`moe_prop()` calculates the uncertainty for the proportion. If the proportion is multiplied by 100, its margin of error must also be multiplied by 100.

Other useful functions include:

- `moe_sum()` for adding estimates
- `moe_prop()` for one estimate as a share of another
- `moe_ratio()` for a ratio where the numerator is not a subset of the denominator

## Coefficient of Variation

The coefficient of variation, or CV, measures the sampling error relative to the estimate:

$$
CV = \frac{SE}{Estimate} \times 100
$$

Because the ACS generally uses a 90% margin of error:

$$
SE = \frac{MOE}{1.645}
$$

Therefore:

```r
cv = (moe / 1.645) / estimate * 100
```

A lower CV means that the margin of error is small relative to the estimate. A higher CV means more sampling uncertainty.

## Change Over Time

ACS 5-year estimates cover overlapping periods. For example:

- The 2022 ACS 5-year estimate covers 2018–2022
- The 2023 ACS 5-year estimate covers 2019–2023

These periods overlap by four years. For a cleaner comparison, use non-overlapping periods such as 2014–2018 and 2019–2023.

When comparing income across time, adjust for inflation before interpreting the difference.

## Creating a Reusable Function

An R function is a reusable set of instructions. This function takes a year as an argument, downloads the data for that year, and creates a period label.

```{r}
#| eval: false

get_poverty <- function(yr) {
  get_acs(
    geography = "county",
    variables = "B07012_002",
    state = "NY",
    year = yr,
    survey = "acs5"
  ) %>%
    mutate(period = paste0(yr - 4, "-", yr))
}
```

Defining the function does not run it. It creates a recipe that can be used later.

```{r}
#| eval: false

poverty_change <- bind_rows(
  get_poverty(2018),
  get_poverty(2023)
)
```

The 2018 data represents 2014–2018, and the 2023 data represents 2019–2023.

## Converting Long Data to Wide Data

The data returned by the two function calls is long:

| County | Period | Estimate |
|---|---|---:|
| Albany County | 2014–2018 | 10,000 |
| Albany County | 2019–2023 | 12,000 |

We can convert it to wide format so the two periods appear in the same row:

```{r}
#| eval: false

poverty_wide <- poverty_change %>%
  select(GEOID, NAME, period, estimate, moe) %>%
  pivot_wider(
    names_from = period,
    values_from = c(estimate, moe)
  )
```

Wide data is useful for comparing two periods because both estimates and both margins of error are side by side.

## Calculating Change and Its Uncertainty

```{r}
#| eval: false

poverty_tested <- poverty_wide %>%
  mutate(
    change = `estimate_2019-2023` - `estimate_2014-2018`,
    se_1 = `moe_2014-2018` / 1.645,
    se_2 = `moe_2019-2023` / 1.645,
    se_diff = sqrt(se_1^2 + se_2^2),
    moe_diff = se_diff * 1.645,
    z = change / se_diff,
    significant = abs(z) > 1.645
  )
```

The difference between two estimates also has uncertainty. We calculate the standard error of the difference, then use it to calculate a margin of error for the change.

`abs(z) > 1.645` tests whether the change is statistically significant using a two-sided 90% confidence level.

```{r}
#| eval: false

poverty_tested %>%
  count(significant)
```

If `significant` is `FALSE`, the estimates may look different but the difference is not large enough to distinguish from sampling variation at the selected confidence level.

## Visualizing Significant Change

```{r}
#| eval: false

poverty_tested %>%
  filter(significant) %>%
  mutate(NAME = str_remove(NAME, " County, New York")) %>%
  ggplot(aes(x = reorder(NAME, change), y = change)) +
  geom_col(aes(fill = change > 0)) +
  geom_errorbar(
    aes(ymin = change - moe_diff, ymax = change + moe_diff),
    width = 0.3
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Statistically Significant Change in Poverty Count",
    subtitle = "2014–2018 versus 2019–2023, ACS 5-year estimates",
    x = NULL,
    y = "Change in people below poverty level"
  )
```

The horizontal line at zero helps us see whether the change is positive or negative.

## Creating a Reliability Table

The `gt` package can create a formatted table with reliability indicators.

```{r}
#| eval: false

reliability_table <- county_data %>%
  mutate(
    cv = (moe / 1.645) / estimate * 100,
    dot = case_when(
      cv < 12  ~ "🟢",
      cv <= 40 ~ "🟡",
      TRUE     ~ "🔴"
    ),
    flag = case_when(
      cv < 12  ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE     ~ "Unreliable"
    ),
    NAME = str_remove(NAME, " County, New York")
  ) %>%
  arrange(desc(cv)) %>%
  select(dot, NAME, estimate, moe, cv, flag)
```

```{r}
#| eval: false

reliability_table %>%
  gt() %>%
  fmt_number(columns = c(estimate, moe), decimals = 0) %>%
  fmt_number(columns = cv, decimals = 1) %>%
  cols_label(
    dot = "",
    NAME = "County",
    estimate = "Estimate",
    moe = "MOE",
    cv = "CV (%)",
    flag = "Reliability"
  ) %>%
  cols_align(align = "center", columns = dot) %>%
  tab_header(
    title = "Population below 100% of the poverty level",
    subtitle = "New York counties, ACS 2023 5-year estimates"
  ) %>%
  tab_source_note(
    "Reliability thresholds: CV < 12% reliable, 12–40% somewhat reliable, and > 40% unreliable."
  )
```

## Key Lessons

This week connected visualization with statistical reasoning.

- Choose a plot that matches the question.
- Use the actual data frame name in `ggplot()`.
- Put variable mappings inside `aes()`.
- Put fixed visual settings outside `aes()`.
- Use error bars to show uncertainty.
- Use `moe_sum()` when adding estimates.
- Use `moe_prop()` when calculating a proportion.
- Avoid overlapping ACS periods when studying change.
- Calculate the uncertainty around differences over time.
- Do not describe a change as meaningful without checking its uncertainty.

## Questions for Further Practice

- Can I create a plot using my own ACS data?
- Can I explain the difference between an estimate, an MOE, a standard error, and a CV?
- Can I explain why two estimates may look different without representing a statistically significant change?
- Can I write a function that pulls data for more than one year?
- Can I create a table that includes estimates, margins of error, and reliability categories?
