---
title: "Week 3: Visualizing Uncertainty and Change"
description: "Using ggplot2, margins of error, and ACS data to study uncertainty over time."
date: 2026-09-11

execute:
  eval: true
  echo: false
  warning: false
  message: false
---

This week connected data visualization with uncertainty in American Community Survey estimates. We practiced using `ggplot2`, calculating margins of error for derived values, comparing ACS estimates across time, and creating reusable R functions.

The examples use New York county data and the 2023 ACS 5-year survey. The charts are included in the rendered page, while the code is hidden.

## Setup

Install these packages once in the R console, not inside this document:

```r
install.packages(c("tidyverse", "tidycensus", "gt"))
```

```{r}
library(tidyverse)
library(tidycensus)
library(gt)
```

## The Grammar of Graphics

`ggplot2` creates plots by mapping variables in a data frame to visual properties.

| Data variable | Plot feature |
|---|---|
| Estimate | Position on an axis |
| Population | Position on an axis |
| Reliability | Color |
| Margin of error | Size or error bars |

This approach is called the **Grammar of Graphics**. A plot is built from separate parts that work together.

## Creating ACS Data

`B17001_002` represents the number of people below 100% of the poverty level.

```{r}
my_variable <- "B17001_002"
my_state <- "NY"

county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Important columns include:

- `GEOID`: geographic identifier
- `NAME`: county name
- `estimate`: Census estimate
- `moe`: margin of error

## Relative Margin of Error

To compare uncertainty across counties, calculate the margin of error as a percentage of the estimate:

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```

A relative margin of error is useful because the same absolute MOE can have different meanings for different-sized estimates.

## Reliability Categories

```{r}
county_data <- county_data %>%
  mutate(
    reliability = case_when(
      moe_pct < 5  ~ "High confidence",
      moe_pct < 10 ~ "Moderate",
      TRUE         ~ "Low confidence"
    )
  )
```

`case_when()` checks conditions from top to bottom. The first condition that is `TRUE` determines the category. The final `TRUE` catches all remaining values.

## Basic Scatterplot

A basic `ggplot2` plot contains the data, aesthetic mappings, and a geometric object.

```{r}
ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point()
```

The first argument must be the name of an actual data frame. For example, `ggplot(data, ...)` would be incorrect here because `data` is only a placeholder.

## Mapping Versus Decorating

When a visual property depends on a variable, put it inside `aes()`:

```{r}
ggplot(county_data, aes(
  x = estimate,
  y = moe_pct,
  color = reliability
)) +
  geom_point()
```

When every point should have the same color, put the color outside `aes()`:

```{r}
ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point(color = "blue")
```

```r
aes(color = reliability)       # maps a variable
geom_point(color = "blue")     # sets one fixed color
```

## Common Geometries

| Question | Useful geometry |
|---|---|
| How is one variable distributed? | `geom_histogram()` |
| How are two variables related? | `geom_point()` |
| How do groups compare? | `geom_boxplot()` or `geom_col()` |
| How does something change over time? | `geom_line()` |
| How uncertain is an estimate? | `geom_errorbar()` |

## Error Bars

An error bar shows the estimate together with its margin of error:

```{r}
county_data %>%
  ggplot(aes(x = NAME, y = estimate)) +
  geom_col() +
  geom_errorbar(
    aes(
      ymin = estimate - moe,
      ymax = estimate + moe
    )
  )
```

The first `aes()` supplies mappings shared by the layers. The second `aes()` supplies the `ymin` and `ymax` mappings needed specifically by `geom_errorbar()`.

The full chart is difficult to read, so we can keep the 15 counties with the largest relative margins of error:

```{r}
top_15 <- county_data %>%
  arrange(desc(moe_pct)) %>%
  slice_head(n = 15) %>%
  mutate(NAME = str_remove(NAME, " County, New York"))

ggplot(top_15, aes(x = reorder(NAME, estimate), y = estimate)) +
  geom_col(fill = "steelblue") +
  geom_errorbar(
    aes(ymin = estimate - moe, ymax = estimate + moe),
    width = 0.3
  ) +
  coord_flip() +
  theme_minimal() +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Counties with the Largest Margins of Error",
    subtitle = "People below 100% of the poverty level, ACS 2023 5-year estimates",
    x = NULL,
    y = "Population below poverty level"
  )
```

## Facets and Themes

A facet creates a separate panel for each category:

```{r}
ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point() +
  facet_wrap(~reliability) +
  theme_minimal()
```

Themes change the appearance of a plot without changing the data.

## Derived Uncertainty

When we calculate a new value from Census estimates, the new value also has uncertainty. We should not simply add individual margins of error when adding estimates.

```{r}
region <- county_data %>%
  filter(NAME %in% c(
    "Hamilton County, New York",
    "Essex County, New York",
    "Greene County, New York"
  ))

region_total <- sum(region$estimate)
region_total_moe <- moe_sum(region$moe, region$estimate)

region_total
region_total_moe
```

The `tidycensus` functions for derived uncertainty are:

- `moe_sum()` for adding estimates
- `moe_prop()` for one estimate as a share of another
- `moe_ratio()` for a ratio where the numerator is not a subset of the denominator

## Proportions and `moe_prop()`

Suppose we want the proportion of people below poverty out of a total population. Both the numerator and denominator have estimates and margins of error.

```{r}
poverty_share <- get_acs(
  geography = "county",
  variables = c(
    below_poverty = "B07012_002",
    total = "B07012_001"
  ),
  state = "NY",
  year = 2023,
  survey = "acs5",
  output = "wide"
) %>%
  mutate(
    share = below_povertyE / totalE,
    share_moe = moe_prop(
      below_povertyE,
      totalE,
      below_povertyM,
      totalM
    ),
    share_pct = share * 100,
    share_moe_pct = share_moe * 100
  )
```

If the proportion is multiplied by 100, its margin of error must also be multiplied by 100.

## Coefficient of Variation

The coefficient of variation, or CV, measures sampling error relative to the estimate:

$$
CV = \frac{SE}{Estimate} \times 100
$$

Because the ACS generally uses a 90% margin of error:

$$
SE = \frac{MOE}{1.645}
$$

Therefore:

```r
cv = (moe / 1.645) / estimate * 100
```

A lower CV means less uncertainty relative to the estimate. A higher CV means more sampling uncertainty.

## Change Over Time

ACS 5-year estimates cover overlapping periods:

- The 2022 ACS 5-year estimate covers 2018–2022
- The 2023 ACS 5-year estimate covers 2019–2023

For a cleaner comparison, use non-overlapping periods such as 2014–2018 and 2019–2023.

When comparing income across time, adjust for inflation before interpreting the difference.

## Creating a Reusable Function

An R function is a reusable set of instructions. This function takes a year as an argument, downloads the data, and creates a period label.

```{r}
get_poverty <- function(yr) {
  get_acs(
    geography = "county",
    variables = "B07012_002",
    state = "NY",
    year = yr,
    survey = "acs5"
  ) %>%
    mutate(period = paste0(yr - 4, "-", yr))
}
```

The function does not run when it is defined. It creates a recipe that can be used later.

```{r}
poverty_change <- bind_rows(
  get_poverty(2018),
  get_poverty(2023)
)
```

## Converting Long Data to Wide Data

The repeated ACS results are initially in long format:

| County | Period | Estimate |
|---|---|---:|
| Albany County | 2014–2018 | 10,000 |
| Albany County | 2019–2023 | 12,000 |

We can convert them to wide format so both periods appear in the same row:

```{r}
poverty_wide <- poverty_change %>%
  select(GEOID, NAME, period, estimate, moe) %>%
  pivot_wider(
    names_from = period,
    values_from = c(estimate, moe)
  )
```

## Calculating Change and Its Uncertainty

```{r}
poverty_tested <- poverty_wide %>%
  mutate(
    change = `estimate_2019-2023` - `estimate_2014-2018`,
    se_1 = `moe_2014-2018` / 1.645,
    se_2 = `moe_2019-2023` / 1.645,
    se_diff = sqrt(se_1^2 + se_2^2),
    moe_diff = se_diff * 1.645,
    z = change / se_diff,
    significant = abs(z) > 1.645
  )
```

The difference between two estimates also has uncertainty. We calculate the standard error of the difference and use it to calculate a margin of error for the change.

## Visualizing Significant Change

```{r}
poverty_tested %>%
  filter(significant) %>%
  mutate(NAME = str_remove(NAME, " County, New York")) %>%
  ggplot(aes(x = reorder(NAME, change), y = change)) +
  geom_col(aes(fill = change > 0)) +
  geom_errorbar(
    aes(ymin = change - moe_diff, ymax = change + moe_diff),
    width = 0.3
  ) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  coord_flip() +
  theme_minimal() +
  labs(
    title = "Statistically Significant Change in Poverty Count",
    subtitle = "2014–2018 versus 2019–2023, ACS 5-year estimates",
    x = NULL,
    y = "Change in people below poverty level"
  )
```

## Creating a Reliability Table

```{r}
reliability_table <- county_data %>%
  mutate(
    cv = (moe / 1.645) / estimate * 100,
    dot = case_when(
      cv < 12  ~ "🟢",
      cv <= 40 ~ "🟡",
      TRUE     ~ "🔴"
    ),
    flag = case_when(
      cv < 12  ~ "Reliable",
      cv <= 40 ~ "Somewhat reliable",
      TRUE     ~ "Unreliable"
    ),
    NAME = str_remove(NAME, " County, New York")
  ) %>%
  arrange(desc(cv)) %>%
  select(dot, NAME, estimate, moe, cv, flag)

reliability_table %>%
  gt() %>%
  fmt_number(columns = c(estimate, moe), decimals = 0) %>%
  fmt_number(columns = cv, decimals = 1) %>%
  cols_label(
    dot = "",
    NAME = "County",
    estimate = "Estimate",
    moe = "MOE",
    cv = "CV (%)",
    flag = "Reliability"
  ) %>%
  cols_align(align = "center", columns = dot) %>%
  tab_header(
    title = "Population below 100% of the poverty level",
    subtitle = "New York counties, ACS 2023 5-year estimates"
  ) %>%
  tab_source_note(
    "Reliability thresholds: CV < 12% reliable, 12–40% somewhat reliable, and > 40% unreliable."
  )
```

## Key Lessons

This week connected visualization with statistical reasoning.

- Choose a plot that matches the question.
- Use the actual data frame name in `ggplot()`.
- Put variable mappings inside `aes()`.
- Put fixed visual settings outside `aes()`.
- Use error bars to show uncertainty.
- Use `moe_sum()` when adding estimates.
- Use `moe_prop()` when calculating a proportion.
- Avoid overlapping ACS periods when studying change.
- Calculate the uncertainty around differences over time.
- Do not describe a change as meaningful without checking its uncertainty.

## Questions for Further Practice

- Can I create a plot using my own ACS data?
- Can I explain the difference between an estimate, an MOE, a standard error, and a CV?
- Can I explain why two estimates may look different without representing a statistically significant change?
- Can I write a function that pulls data for more than one year?
- Can I create a table that includes estimates, margins of error, and reliability categories?
