---
title: "Week 3: Visualizing Uncertainty and Change"
description: "Using ggplot2, margins of error, and ACS data to study uncertainty over time."
date: 2026-09-11
---

## What We Covered

This week focused on:

- Creating plots with `ggplot2`
- Measuring uncertainty in ACS estimates
- Calculating uncertainty for derived values
- Comparing ACS estimates across time
- Writing reusable R functions
- Presenting results clearly in tables and figures

## The Grammar of Graphics

`ggplot2` is based on the idea that a plot is created by connecting variables in a data frame to visual properties.

| Data variable | Plot feature |
|---|---|
| Estimate | Position on an axis |
| Population | Position on an axis |
| Reliability | Color |
| Margin of error | Size or error bars |

This idea is called the **Grammar of Graphics**. A plot is built from separate parts that work together.

## The Basic Parts of a ggplot

A basic plot usually contains:

1. Data
2. Aesthetic mappings
3. A geometric object

```{r}
#| eval: false

ggplot(county_wide, aes(x = popE, y = moe_pct)) +
  geom_point()
```

`ggplot(county_wide)` creates the plotting canvas using `county_wide`.

`aes()` maps variables from the data to visual properties.

`geom_point()` tells R to display the observations as points.

The data frame must be named correctly. This will cause an error:

```{r}
#| eval: false

ggplot(data, aes(x = population, y = moe_pct)) +
  geom_point()
```

Here, `data` is only a placeholder. It is not the name of an actual data frame.

A correct version uses the real object and column names:

```{r}
#| eval: false

ggplot(county_wide, aes(x = popE, y = moe_pct)) +
  geom_point()
```

## Mapping Versus Decorating

If a visual property depends on a variable, place it inside `aes()`:

```{r}
#| eval: false

ggplot(county_data, aes(
  x = estimate,
  y = moe_pct,
  color = reliability
)) +
  geom_point()
```

The points receive different colors based on the `reliability` column.

If every point should have the same color, place the color outside `aes()`:

```{r}
#| eval: false

ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point(color = "blue")
```

All points are blue in this example.

```r
aes(color = reliability)       # maps a variable
geom_point(color = "blue")     # sets one fixed color
```

## Common Geometries

The geometry, or `geom`, determines the type of chart.

| Question | Useful geometry |
|---|---|
| How is one variable distributed? | `geom_histogram()` |
| How are two variables related? | `geom_point()` |
| How do groups compare? | `geom_boxplot()` or `geom_col()` |
| How does something change over time? | `geom_line()` |
| How uncertain is an estimate? | `geom_errorbar()` |

For example, a histogram shows how values are distributed:

```{r}
#| eval: false

ggplot(county_data, aes(x = moe_pct)) +
  geom_histogram()
```

A scatterplot shows the relationship between two variables:

```{r}
#| eval: false

ggplot(county_wide, aes(x = popE, y = moe_pct)) +
  geom_point()
```

A bar chart compares categories:

```{r}
#| eval: false

ggplot(summary_data, aes(x = reliability, y = n_counties)) +
  geom_col()
```

An error bar can show an estimate together with its margin of error:

```{r}
#| eval: false

ggplot(county_data, aes(x = NAME, y = estimate)) +
  geom_point() +
  geom_errorbar(
    aes(
      ymin = estimate - moe,
      ymax = estimate + moe
    )
  )
```

## Facets and Small Multiple Plots

A facet creates a separate panel for each category of a variable.

```{r}
#| eval: false

ggplot(county_data, aes(x = estimate, y = moe_pct)) +
  geom_point() +
  facet_wrap(~reliability)
```

This creates one panel for each reliability category.

Facets are useful when groups would be difficult to compare in a single plot.

## Themes and Labels

A theme changes the appearance of a plot. It does not change the data.

```{r}
#| eval: false

ggplot(county_wide, aes(x = popE, y = moe_pct)) +
  geom_point() +
  theme_minimal()
```

Other useful themes include:

```r
theme_bw()
theme_classic()
theme_light()
```

Labels make a plot easier to understand:

```{r}
#| eval: false

ggplot(county_wide, aes(x = popE, y = moe_pct)) +
  geom_point() +
  labs(
    title = "Population and Relative Margin of Error",
    x = "County population",
    y = "Margin of error percentage"
  ) +
  theme_minimal()
```

When county names are difficult to read, a horizontal bar chart may work better:

```{r}
#| eval: false

ggplot(county_data, aes(x = NAME, y = estimate)) +
  geom_col() +
  coord_flip()
```

## Coefficient of Variation

The coefficient of variation, or CV, measures the size of the sampling error relative to the estimate.

$$
CV = \frac{SE}{Estimate} \times 100
$$

The ACS generally reports a 90% margin of error. To estimate the standard error:

$$
SE = \frac{MOE}{1.645}
$$

Therefore:

```r
cv = (moe / 1.645) / estimate * 100
```

A lower CV means the margin of error is small compared with the estimate. A higher CV means the estimate has more sampling uncertainty.

The class used these general categories:

| CV | Interpretation |
|---|---|
| Less than 15% | High reliability |
| 15% to 30% | Medium reliability |
| Greater than 30% | High sampling error |

These are guidelines for interpreting the data. They are not universal laws.

## Derived Uncertainty

When we calculate a new value from Census estimates, the new value also has uncertainty.

For example:

$$
Poverty\ Rate =
\frac{People\ in\ poverty}{Total\ population}
\times 100
$$

Both the numerator and denominator have their own margins of error. We cannot simply use the original margin of error for the new percentage.

The `tidycensus` package provides functions for derived margins of error:

- `moe_sum()` calculates uncertainty for a sum
- `moe_prop()` calculates uncertainty for a proportion
- `moe_ratio()` calculates uncertainty for a ratio

If we add estimates from several counties, we should not simply add their margins of error. We should use `moe_sum()`.

If we calculate the share of people in poverty out of the total population, we should use `moe_prop()`.

The important idea is that a derived estimate needs its own margin of error.

## Studying Change Over Time

ACS 5-year estimates cover overlapping periods.

For example:

- The 2022 ACS 5-year estimate covers 2018–2022
- The 2023 ACS 5-year estimate covers 2019–2023

These periods overlap by four years. This makes it difficult to interpret the change because much of the data comes from the same years.

For a cleaner comparison, we can use non-overlapping periods such as:

- 2014–2018
- 2019–2023

When comparing income across time, we should also account for inflation. An income value from an earlier period may need to be converted into current dollars before comparison.

## Difference Between Two Estimates

Suppose we have these estimates:

- 2014–2018: `$45,000 ± $8,000`
- 2019–2023: `$62,000 ± $11,000`

The difference is:

```text
$62,000 - $45,000 = $17,000
```

However, `$17,000` is not an exact value because both estimates contain uncertainty.

First, convert each margin of error into a standard error:

```text
SE1 = 8,000 / 1.645
SE2 = 11,000 / 1.645
```

Then calculate the standard error of the difference:

$$
SE_{diff} =
\sqrt{SE_1^2 + SE_2^2}
$$

Finally, calculate the confidence interval:

$$
Difference \pm (1.645 \times SE_{diff})
$$

This tells us whether the observed change is large enough to be distinguished from sampling variation.

## Testing Whether Change Is Significant

The change can also be expressed as a z-score:

$$
z =
\frac{Estimate_2 - Estimate_1}{SE_{diff}}
$$

For a two-sided 90% confidence test:

```r
abs(z) > 1.645
```

If this returns `TRUE`, the change is statistically significant at the 90% confidence level.

If it returns `FALSE`, we cannot confidently say that the difference represents a real change rather than sampling variation.

Statistical significance does not prove that one event caused the change. It only tells us whether the difference is large compared with the estimated uncertainty.

## Writing Reusable Functions

An R function is a reusable set of instructions.

```{r}
#| eval: false

get_poverty <- function(year) {
  get_acs(
    geography = "county",
    variables = c(
      poverty = "B17001_002",
      total = "B17001_001"
    ),
    state = "NY",
    year = year,
    survey = "acs5"
  ) %>%
    mutate(
      period = paste0(year - 4, "-", year)
    )
}
```

Defining the function does not run it. It creates a recipe that can be used later.

We can run the function for different years:

```{r}
#| eval: false

data_2018 <- get_poverty(2018)
data_2023 <- get_poverty(2023)
```

We can also use `purrr` to run it for several years:

```{r}
#| eval: false

library(purrr)

poverty_data <- map_dfr(
  c(2018, 2023),
  get_poverty
)
```

`map_dfr()` runs the function for each year and combines the results into one data frame.

## Long and Wide Data

Repeated ACS pulls are usually returned in long format.

Long data may look like this:

| County | Period | Estimate |
|---|---|---:|
| Albany County | 2014–2018 | 10,000 |
| Albany County | 2019–2023 | 12,000 |

Wide data places the time periods side by side:

| County | Estimate old | Estimate new |
|---|---:|---:|
| Albany County | 10,000 | 12,000 |

Wide data is often easier for calculating change because both periods are in the same row.

The `pivot_wider()` function can convert long data into wide data.

## Presenting Results with Tables

The `gt` package can create polished tables.

A table can include:

- The estimate
- The margin of error
- The coefficient of variation
- A reliability category
- A confidence interval
- A short interpretation

The goal is to make the result understandable to someone who may not know the technical details.

## Hiding Code in Quarto

The code can be hidden in the rendered HTML while the results remain visible.

In the YAML header, we can use:

```yaml
execute:
  echo: false
  message: false
  warning: false
```

`echo: false` hides the code.

`message: false` hides messages produced while the code runs.

`warning: false` hides warnings. Warnings should still be checked during development before hiding them in the final report.

For these study notes, example chunks use:

```r
#| eval: false
```

This means Quarto displays the code but does not run it. This is useful when the code refers to example objects that have not been created in the document.

## Week 3 Assignment

The assignment asks us to examine a claim about change over time using ACS data.

The general workflow is:

1. Find an article or example that discusses change using ACS data.
2. Identify the variable and geographic areas being studied.
3. Determine whether the ACS periods overlap.
4. Pull non-overlapping ACS estimates when possible.
5. Calculate the estimate and its margin of error.
6. Calculate the coefficient of variation.
7. Calculate the change between the two periods.
8. Calculate the uncertainty around the change.
9. Test whether the change is statistically significant.
10. Decide whether the original claim is supported.
11. Present the results in a clear HTML document.

For a poverty rate, the numerator and denominator both have uncertainty, so `moe_prop()` should be used to calculate the margin of error for the rate.

The analysis should not only report the difference between two estimates. It should also explain how certain that difference is.

## Key Lessons

This week connected visualization with statistical reasoning.

A plot is not just decoration. The variables mapped to its axes, colors, sizes, and shapes affect how the reader understands the data.

ACS estimates also need to be interpreted with their margins of error. A large difference between two estimates does not automatically mean that a real change occurred.

The main lessons were:

- Choose a plot that matches the question.
- Use the real data frame name in `ggplot()`.
- Separate variable mappings from fixed decorations.
- Use CV to compare uncertainty across estimates.
- Use the correct `tidycensus` function for derived margins of error.
- Avoid comparing overlapping ACS periods when studying change.
- Calculate the uncertainty around differences over time.
- Explain whether an observed change is statistically significant.
- Present the result clearly for technical and non-technical readers.

## Questions for Further Practice

- Can I create a plot using my own ACS data?
- Can I explain the difference between an estimate, an MOE, a standard error, and a CV?
- Can I explain why two estimates may look different without representing a statistically significant change?
- Can I write a function that pulls data for more than one year?
- Can I create a table that includes estimates, margins of error, and reliability categories?
