---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Frederick Abalo"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---



## Setup

```{r}
library(tidyverse)
library(tidycensus)
```

---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.

::: {.callout-tip}
## Delete this box

Rewrite the paragraph below in your own words once you've picked your state and
variable. Two or three sentences saying what you're looking at and why.
:::

**In this analysis, I look at [YOUR VARIABLE] across counties in [YOUR STATE], and
ask which communities we measure most and least reliably.**

---

## Choosing what to measure

::: {.callout-note}
## Instructions — delete this box

Pick **a state** and **a variable**. Anything except Pennsylvania and median
household income, which we did in class.

Some variables that work well:

| Code | What it is |
|------|-----------|
| `B19013_001` | Median household income |
| `B25064_001` | Median gross rent |
| `B25077_001` | Median home value |
| `B01002_001` | Median age |
| `B08303_001` | Total commuters (travel time table) |
| `B15003_022` | Number with a bachelor's degree |

Want something else? Run this in your **console** (not in a chunk):

```r
vars <- load_variables(2023, "acs5")
View(vars)
```

Then use the search box.
:::

```{r}
my_state <- "CA"        # two-letter abbreviation, in quotes
my_variable <- "B01002_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```



California has 58 counties and that number matches the number of rows (58) returned by checking the dimensions of the data.
---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```



```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```
---

Alpine County is the least reliable county in the dataset, with the highest relative margin of error(moe_pct) at approximately 11.68%.

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 1  ~ "High confidence",
    moe_pct < 4  ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

Using the original thresholds, 56 counties were classified as high confidence, 1 as moderate, and 1 as low confidence. Because this classification was highly concentrated in one category, I adjusted the thresholds to 1% and 4% to make differences in relative margin of error more informative for this county-level dataset.

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```



```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 5  ~ "High confidence",
    moe_pct < 15 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```
The high-confidence counties had an average population of 700,685, compared with 2,213 for the moderate-confidence counties. Although no counties were classified as low confidence, the pattern suggests that smaller counties tend to have larger relative margins of error and less reliable estimates.
---

## What this means for policy

::: {.callout-important}
## Instructions — delete this box

**Write 200–250 words.** 

Address all three:

1. **Which communities in your state do we know the least about?** Name specific
   counties from your results.
2. **Why?** Connect the pattern to how the ACS actually works — it's a sample, and
   smaller places yield smaller samples.
3. **What could go wrong?** Imagine a state agency allocating funding using your
   variable, treating every county's number as equally certain. Who is likely to be
   hurt, and how?

Use your own numbers. 
:::

**[Your 200–250 word reflection goes here.]**

---

## Optional challenges - For those craving more

::: {.callout-note}
## Delete this box

**None of these are graded.** There are no extra points. A complete, well-written assignment earns a 2 either way.

They're here for anyone who finished the above and found it straightforward. Pick
one (or more) that interests you. 

1. **Propagate the margin of error.** If you compute a derived value — a percentage,
   a ratio, a sum — its uncertainty depends on the uncertainty of its inputs, and
   you can't just carry the original MOE along. The Census Bureau publishes formulas
   for this, and `tidycensus` implements them: `moe_sum()`, `moe_prop()`,
   `moe_ratio()`. Compute something derived, and get its margin of error right.

2. **Parameterize the report.** Add `params:` to the YAML so the state is an input
   rather than hard-coded, and the same file regenerates for any state. Render it
   from the terminal with `quarto render assignment2.qmd -P state:NY`.

3. **Write a function and iterate.** Wrap the data pull in a function, then use
   `purrr::map_dfr()` to run it across several states at once. Does the
   small-places-are-noisier pattern hold everywhere, or are some states different?

4. **Change the geographic scale.** Run the same variable at `geography = "tract"`
   inside one county. Compare the margins of error to what you saw at county level.
   What does that imply about the resolution at which you can actually trust ACS
   data?

If you try one, add a short section describing what you did and what you found.
:::

---

## Before you submit

::: {.callout-note}
## Delete this box — checklist

- [ ] Your name is in the YAML header
- [ ] Every instruction callout is deleted
- [ ] Every `[bracketed prompt]` is replaced with your own writing
- [ ] No `___` blanks remain anywhere
- [ ] The file is at `assignments/assignment2.qmd`
- [ ] **Session → Restart R**, then render the whole project. It works.
- [ ] Committed and pushed
- [ ] You opened your live site and looked at the published page
:::
