---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Frederick Abalo"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---

## Setup

```{r}
library(tidyverse)
library(tidycensus)
```


## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.


## Choosing what to measure

```{r}
my_state <- "CA"        # two-letter abbreviation, in quotes
my_variable <- "B01002_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```


California has 58 counties and that number matches the number of rows (58) returned by checking the dimensions of the data.

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```



```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```


Alpine County is the least reliable county in the dataset, with the highest relative margin of error(moe_pct) at approximately 11.68%.


## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 1  ~ "High confidence",
    moe_pct < 4  ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```

Using the original thresholds, 56 counties were classified as high confidence, 1 as moderate, and 1 as low confidence. Because this classification was highly concentrated in one category, I adjusted the thresholds to 1% and 4% to make differences in relative margin of error more informative for this county-level dataset.



## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```



```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 1  ~ "High confidence",
    moe_pct < 4 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```
The high-confidence counties had an average population of 924,758, compared with 31,394 for the moderate-confidence counties and only 6,560.25 for low-confidence counties. The pattern suggests that smaller counties tend to have larger relative margins of error and less reliable estimates.

## What this means for policy

The least reliable estimates came from Alpine, Sierra, Mono, and Modoc Counties. Their average population was 6,560, compared with 924,758 for the 42 high-confidence counties.

This happens because the ACS uses a sample of households. Smaller counties have fewer households in the sample, so their estimates have more uncertainty and larger margins of error.

If a state agency treated every estimate as equally reliable, these rural counties could receive less funding than they need for healthcare, transportation, housing, or services for older residents. Funding decisions should account for both the estimate and its margin of error.

## My Reflections

This week helped me understand how R can be used to move from raw data to a clear result. I practiced using tidyverse functions such as `filter()`, `select()`, `mutate()`, `arrange()`, `group_by()`, and `summarize()`. Using pipes also made the code easier to follow because I could connect several steps without creating a new object after every operation.

I also learned how to use `tidycensus` to download American Community Survey data. Before this, I mostly thought of Census data as a table of numbers. Working with estimates and margins of error showed me that the numbers also have different levels of reliability. In my county analysis, the smaller counties had much larger relative margins of error than the larger counties. The four low-confidence counties had an average population of 6,560, while the 42 high-confidence counties had an average population of 924,758.

The exercise also showed me why it is important to understand the structure of the data before analyzing it. I had to distinguish between long and wide formats, identify the estimate and margin-of-error columns, and make sure my reliability cutoffs were applied consistently. I am still getting used to R syntax, but the workflow is becoming clearer. The optional, non-scoring problems will require more practice because I need to build a stronger understanding of the concepts before I can work through them independently.


